Type Refinements for the Working Class

### 2015-11-14 paper.tex
[paper] special arrow for display
explain "internal judgments" in the topos logic
It's impredicative, but this is a mere formalization of what goes on in
the meaning explanations all the way back to at least Heyting. For
instance, to know "P prop" is to furnish another judgment "P verified"
and its meaning, which is in general an impredicative construction.

I think that what Mark van Atten has said about impredicativity is
relevant, which is that predicativity is not an intrinsic constraint on
constructions by the creating subject, but is rather something that may
be observed *about* construc

\newcommand\EvalN[5]{\IMode{#1}\parallel\IMode{#4}\Downarrow^{\OMode{#3}}_{\IMode{#2}}\OMode{#5}}
\newcommand\Exprs{\mathcal{E}}
\newcommand\Values{\mathcal{V}}
\newcommand\Naturals{\mathbb{N}}
\newcommand\NaturalsPsh{\mathcal{N}}

\tikzset{
   commutative diagrams/.cd,
   arrow style = tikz,
   diagrams={>=latex}
}

@@

For any abt signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we can deploy
a generic logical framework. Recall from~\cite{sterling-morrison:2015} that
$\SCtx$ is the category of symbol-contexts and their injective, sort-preserving
renamings. Then, fix a small category $\mathcal{W}$ of worlds along with a map
$\Of{\pi}{\mathcal{W}\ra\SCtx}$ such that $\pi$ is a Grothendieck
fibration~\cite{jacobs:2001, taylor:1999}:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&, row sep = large]
    {\IMode{\mathcal{W}}}
      \arrow[dash, d]
\&
    {\IMode{{w}_{i}}}
      \arrow[drrr]
      \arrow[dotted, dr]
      \arrow[mapsto, dd]
\&
    \null
\&
    \null
\&
    \null
\\
    {\IMode{{\pi}}}
      \arrow[d]
\&
    \null
\&
    {\OMode{{w}_{j}}}
      \arrow[rr, swap, "\OMode{{\iota}_{\varrho}}"]
      \arrow[mapsto, dd, bend right = 50]
      \arrow[mapsto, d]
\&
    \null
\&
    |[alias = Wk]|{\IMode{{w}_{k}}}
      \arrow[mapsto, d]
\\
    {\IMode{\SCtx}}
\&
    |[alias = Ui]|{\IMode{{\Upsilon}_{i}}}
      \arrow[dr]
\&
    {\OMode{\scriptstyle{{\pi}({w}_{j})}}}
      \arrow[rr, "\OMode{{\pi}({\iota}_{\varrho})}", ""{name = Msrc, swap}]
      \arrow[equals, d]
\&
    \null
\&
    {\IMode{\scriptstyle{{\pi}({w}_{k})}}}
      \arrow[equals, d]
\\
    \null
\&
    \null
\&
    {\IMode{{\Upsilon}_{j}}}
      \arrow[rr, swap, "{\IMode{\varrho}}", ""{name = Mtgt, swap}]
\&
    \null
\&
    |[alias = Uk]|{\IMode{{\Upsilon}_{k}}}
      \arrow[mapsto, from = Wk, to = Uk, bend right = 50, crossing over]
      \arrow[equals, from = Msrc, to = Mtgt]
      \arrow[from = Ui, to = Uk, crossing over]
  \end{tikzcd}
\]
Then, a \emph{judgment} shall be interpreted as an object in the presheaf
category $\Define{\mathbb{J}}{\Sets^{\OpCat{\mathcal{W}}}}$; in other words, a
judgment is identified with the intensional set of its renaming-sensitive
derivations.

@@

  In other words, the hypothetical judgment
  $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ is evident when there shall forevermore
  be an effective transformation of evidence for the antecedent into evidence
  for the consequent. It is important to keep in mind that this is a
  \emph{semantic} consequence and therefore expresses admissibilities, in
  contrast to the \emph{logical} consequence which expresses
  derivabilities~\cite{harper:2016}. We will not have any need for the latter.
\end{definition}

@@

@@

In this section we generalize Howe's notion of lazy computation
system~\cite{howe:1989} to the multi-sorted, symbol-parameterized setting. An
\emph{lazy computation language} is an abt signature
$\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$ along with a distinguished copresheaf
$\Of{\CanOperators}{\Sets^{\SCtx\times\Discrete{\Arities}}}$ of
\emph{canonical} operators such that $\IsSubsetEq{\CanOperators}{\Operators}$.
For a lazy computation language
$\Match{L}{\Tuple{\Sorts,\Operators,\CanOperators}}$, let us define the
copresheaves of closed expressions and closed values over symbol contexts and sorts as
follows:
\begin{align*}
  \ADefine{\Exprs_\tau(\Upsilon)}{
    \MkSet{M \mid \IsAbtUnmoded{\cdot}{\Upsilon}{\cdot}{M}{\tau}}
  }\\
  \ADefine{\Values_\tau(\Upsilon)}{
    \MkSet{
      M\equiv\App{\vartheta}{\vec{E}}
      \mid M\in\Exprs_\tau(\Upsilon)\land \exists a. \CanOperators\Pair{\Upsilon}{a}\ni\vartheta
    }
  }
\end{align*}

Then, a \emph{lazy computation system} (lcs) is a lazy computation language
$\Match{L}{\Pair{\Sigma}{\CanOperators}}$ along with an indexed
$\equiv_\alpha$-functional relation
$\Of{\Downarrow}{\Pow{\Exprs\times\Values\times\NaturalsPsh}}$ where
$\NaturalsPsh$ is the constant presheaf $\MapsTo{(\Upsilon, \tau)}{\Naturals}$.
We will write $\EvalN{\Upsilon}{\tau}{n}{M}{N}$ when
$\Member{\Tuple{M,N,n}}{\Downarrow_\tau(\Upsilon)}$, to mean that $M$ evaluates
to $N$ in $n$ steps.

>>>>

\newcommand\EvalN[5]{\IMode{#1}\parallel\IMode{#4}\Downarrow^{\OMode{#3}}_{\IMode{#2}}\OMode{#5}}
\newcommand\Eval[4]{\EvalN{#1}{#2}{}{#3}{#4}}
\newcommand\Exprs{\mathcal{E}}
\newcommand\Values{\mathcal{V}}
\newcommand\Naturals{\mathbb{N}}
\newcommand\NaturalsPsh{\mathcal{N}}

\tikzset{
  commutative diagrams/.cd,
  arrow style = tikz,
  diagrams = {>=latex},
}
\tikzstyle{displays} = [
  ->,
  >=open triangle 45,
]

@@

For any abt signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we can deploy
a generic logical framework. Recall from~\cite{sterling-morrison:2015} that
$\SCtx$ is the category of symbol-contexts and their injective, sort-preserving
renamings. Then, fix a small category $\mathcal{W}$ of worlds along with a map
$\Of{\pi}{\mathcal{W}\ra\SCtx}$ such that $\pi$ is a Grothendieck
fibration~\cite{jacobs:2001, taylor:1999}:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&, row sep = large]
    {\IMode{\mathcal{W}}}
      \arrow[dash, d]
\&
    {\IMode{{w}_{i}}}
      \arrow[drrr]
      \arrow[dotted, dr]
      \arrow[mapsto, dd]
\&
    \null
\&
    \null
\&
    \null
\\
    {\IMode{{\pi}}}
      \arrow[displays, d]
\&
    \null
\&
    {\OMode{{w}_{j}}}
      \arrow[rr, swap, "\OMode{{\iota}_{\varrho}}"]
      \arrow[mapsto, dd, bend right = 50]
      \arrow[mapsto, d]
\&
    \null
\&
    |[alias = Wk]|{\IMode{{w}_{k}}}
      \arrow[mapsto, d]
\\
    {\IMode{\SCtx}}
\&
    |[alias = Ui]|{\IMode{{\Upsilon}_{i}}}
      \arrow[dr]
\&
    {\OMode{\scriptstyle{{\pi}({w}_{j})}}}
      \arrow[rr, "\OMode{{\pi}({\iota}_{\varrho})}", ""{name = Msrc, swap}]
      \arrow[equals, d]
\&
    \null
\&
    {\IMode{\scriptstyle{{\pi}({w}_{k})}}}
      \arrow[equals, d]
\\
    \null
\&
    \null
\&
    {\IMode{{\Upsilon}_{j}}}
      \arrow[rr, swap, "{\IMode{\varrho}}", ""{name = Mtgt, swap}]
\&
    \null
\&
    |[alias = Uk]|{\IMode{{\Upsilon}_{k}}}
      \arrow[mapsto, from = Wk, to = Uk, bend right = 50, crossing over]
      \arrow[equals, from = Msrc, to = Mtgt]
      \arrow[from = Ui, to = Uk, crossing over]
  \end{tikzcd}
\]
Then, a \emph{judgment} shall be interpreted as an object in the presheaf
category $\Define{\mathbb{J}}{\Sets^{\OpCat{\mathcal{W}}}}$; in other words, a
judgment is identified with the intensional set of its renaming-sensitive
derivations. $\mathbb{J}$, like all presheaf categories, is a topos.

@@

  In other words, the hypothetical judgment
  $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ is evident when there shall forevermore
  be an effective transformation of evidence for the antecedent into evidence
  for the consequent. It is important to keep in mind that this is a
  \emph{semantic} consequence and therefore expresses admissibilities, in
  contrast to the \emph{logical} consequence which expresses
  derivabilities~\cite{harper:2016}. We will not have any need for the latter.

  \paragraph{Notation}
  For a hypothetical judgment with multiple hypotheses, we will write
  $\HypJ{\mathcal{J}_3}{\mathcal{J}_1, \mathcal{J}_2}$ rather than
  $\HypJ{\mathcal{J}_3}{\mathcal{J}_1\times\mathcal{J}_2}$ or
  $\HypJ{(\HypJ{\mathcal{J}_3}{\mathcal{J}_2})}{\mathcal{J}_1}$.
\end{definition}

@@

\begin{definition}[Internal Judgments]
  It often happens that the evidence for a judgment is the knowledge of the
  meaning of a new form of judgment. This can be neatly expressed in our topos
  using the power object $\Pow{-}$: for any judgment
  $\mathcal{J}\in\mathbb{J}$, the derivations of $\Pow{\mathcal{J}}$ shall
  consist in subsets of the derivations of $\mathcal{J}$. The evidence of
  $\Pow{\mathcal{J}}$ shall be called an \emph{internal judgment} $J$. In most
  cases, we will say something like ``To know $\mathcal{J}$ is to have a
  judgment $\mathcal{J'}$'' rather than appealing directly to the power object
  or the subobject classifier.
\end{definition}

@@

In this section we generalize Howe's notion of lazy computation
system~\cite{howe:1989} to the multi-sorted, symbol-parameterized setting. An
\emph{lazy computation language} is an abt signature
$\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$ along with a distinguished copresheaf
$\Of{\CanOperators}{\Sets^{\SCtx\times\Discrete{\Arities}}}$ of
\emph{canonical} operators such that $\IsSubsetEq{\CanOperators}{\Operators}$.
For a lazy computation language
$\Match{L}{\Tuple{\Sorts,\Operators,\CanOperators}}$, let us define the
presheaves of closed expressions and closed values over symbol contexts and sorts as
follows:
\begin{align*}
  \ADefine{\Exprs_\tau(w)}{
    \MkSet{M \mid \IsAbtUnmoded{\cdot}{\pi(w)}{\cdot}{M}{\tau}}
  }\\
  \ADefine{\Values_\tau(w)}{
    \MkSet{
      M\equiv\App{\vartheta}{\vec{E}}
      \mid M\in\Exprs_\tau(w)\land \exists a. \CanOperators\Pair{\pi(w)}{a}\ni\vartheta
    }
  }
\end{align*}

\begin{remark}
  If read as a judgment, $\Exprs_\tau$ (resp.\ $\Values_\tau$) means ``there is
  a closed expression (resp.\ value) of sort $\tau$'', and has as its evidence any closed
  expression (resp.\ value) of sort $\tau$.
\end{remark}

Then, a \emph{lazy computation system} (lcs) is a lazy computation language
$\Match{L}{\Pair{\Sigma}{\CanOperators}}$ along with a
$\equiv_\alpha$-functional evaluation relation. Formally, the evaluation
relation is given in our logical framework by a sort-indexed family of
intensional sets of evaluation rules
$\Of{\mathfrak{R}^\tau_\Downarrow}{\mathbb{J}}$, such that the judgment
$\HypJ{\Pow{\mathfrak{R}^\tau_\Downarrow}}{\Exprs_\tau, \Values_\tau,
\NaturalsPsh}$ shall be evident by means of some internal family of judgments
$[\Downarrow_\tau]$, where $\NaturalsPsh$ is the constant presheaf
$\MapsTo{w}{\Naturals}$. In other words, we shall forevermore know what it
means to know that some expression evaluates to some value in $n$ steps.  By
this token, we write $\EvalN{\Upsilon}{\tau}{n}{M}{N}$ to mean the judgment
whose derivations at $w$ are the set $[\Downarrow_\tau](w,M,N,n)$.
